---
title: "Tidying and Wrangling"
format: html
---

```{r}
#| message: false
library(dplyr)
library(tidyr)
library(readr)
```
# Data Sources

## Read in our data 

```{r}
#| message: false
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

## Explore

```{r}
colnames(catch_original)

head(catch_original)

summary(catch_original)

unique(catch_original$Region)
```

%>% pipe key
Mac: cmd shift M

# Cleaning and Reshaping

## Using the select(function)


```{r}
# select() to keep the columns that we want
catch_data <- catch_original %>% 
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)
head(catch_data)

# use - to drop columns yo dont't want
catch_data <- catch_original %>%  select(-All, -notesRegCode)
head(catch_data)
```
## ckeck column types

```{r}
glimpse(catch_data)
```
## change column contents with mutate()
```{r}
#change chinook to numeric

catch_clean <- catch_data %>% 
  mutate(Chinook = as.numeric(Chinook))
```
```{r}
is.na(catch_clean$Chinook)

# get the location of the NA value
i <- which(is.na(catch_clean$Chinook))
i

catch_data[i,]
catch_clean <- catch_data %>% 
  mutate(Chinook = if_else(condition = Chinook == "I",
                           true = "1",
                           false = Chinook),
         Chinook = as.numeric(Chinook))
head(catch_clean)

```
## Changing file shape using pivot_longer() and pivot_wider()

```{r}
catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year),
               names_to = "Species",
               values_to = "Catch")
head(catch_long)

catch_wide <- catch_long %>% 
  pivot_wider(names_from = Species, 
              values_from = Catch)
```
## rename columns with rename()
```{r}
catch_long <- catch_long %>% 
  rename(catch_thousands = Catch)
head(catch_long)
```
```{r}
catch_long <- catch_long %>% 
  mutate(Catch = catch_thousands * 1000) %>% 
  select(-catch_thousands)

head(catch_long)
```
## summary statistics usign group_by and summarize()
```{r}
# get average catch by region
mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(Catch))
head(mean_region)

# number of rows by each group
n_region <- catch_clean %>% 
  group_by(Region) %>% 
  summarise(n = n())
head(n_region)
```
# Filtering rows using filter()
```{r}
sse_catch <- catch_long %>% 
  filter("Region" == "SSE")

head(sse_catch)

# try filtering where catch is over 1M fidh
catch_million <- catch_long %>% 
  filter(Catch >= 1000000)
head(catch_million)

chinook_sse <- catch_long %>% 
  filter(Region == "SSE" & Species == "Chinook")
head(chinook_sse)
```
## Sorting data with arrange()
```{r}
mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarise(mean_catch = mean(Catch)) %>% 
  arrange(desc(mean_catch)) ## desc() is for descending order 
head(mean_region)
```

## Combining datasets together with join()
```{r}
region_defs <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>% 
    select(code, mgmtArea)

head(region_defs)

catch_joined <- left_join(catch_long, region_defs, by = c("Region" = "code"))
```
```{r}
region_defs <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>% 
    select(code, mgmtArea) %>% 
  rename(Region = code, Region_name = mgmtArea)

catch_joined <- left_join(catch_long, region_defs, by = c("Region"))
head(catch_joined)
```

## split and combine values and columns

```{r}
sites_df <- data.frame(site = c("TLK-001",
                                "TLK-002",
                                "BAR-005"))
head(sites_df)

split_df <- sites_df %>% 
  separate(site, into = c("loc", "site_number"), sep = "-")

head(split_df)

```
```{r}
dates_df <- data.frame(
  year =c("1930", "1930", "1932"),
  month = c("12", "12", "12"),
  day = c("14", "15", "16")
                       )
head(dates_df)
dates_unite <- dates_df %>% 
  unite(col = date, year, month, day, sep = "-")
head(dates_unite)
```
```{r}
mean_region <- catch_original %>% 
  select(-All, -notesRegCode) %>% 
  mutate(Chinook = if_else(Chinook == "I",
                           "1",
                           Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook)) %>% 
  pivot_longer(-c(Region, Year),
               names_to = "Species",
               values_to = "Catch") %>% 
  mutate(Catch = Catch * 1000) %>% 
  group_by(Region) %>% 
  summarise(mean_catch = mean(Catch)) %>% 
  arrange(desc(mean_catch))

head(mean_region)
```
## save our results with write.csv()

```{r}
library(here)
write_csv(mean_region, here("data/mean_catch_by_region.csv"))
```

